---
title: "Analyzing AMZN Stock"
output: pdf_document
---

# Introduction

This document will attempt an organized and clear attempt at predicting Amazon (AMZN) closing stock prices using Time Series Models. In particular, ARTFIMA model.

```{r, echo = F, message = F, warning = F}
library(artfima)
library(quantmod)
# library(stats)   # for arima.sim and acf
```

# Analyze

To begin our analysis, we will plot the raw closing prices from 2000 - 2017. As you can see, there has been fairly explosive growth starting in January 2009
```{r}
getSymbols("AMZN", from = "2000-01-03", to = "2017-01-03")

# Extract adjusted closing prices
amzn_close <- Cl(AMZN)

# Plot closing prices
plot(amzn_close, main = "Amazon Closing Prices (2000–2017)", col = "darkblue")
```

Next, we are interested in examining the squared log returns per day. Shown below. Very spikey.

```{r}
amzn_ret <- diff(log(amzn_close))
amzn_ret <- na.omit(amzn_ret)

amzn_sqr <- amzn_ret^2

# Plot returns
plot(amzn_sqr, main = "Amazon Log Returns (2000–2017)", col = "darkred")
```

Now we check the ACF and PACF to see if we have a clear modeling choice. The ACF seems to be cyclical to some degree, and long. Implying a long time series

```{r}

# adf.test(amzn_ret)
# par(mfrow = c(1,2))
acf(amzn_sqr, lag.max = 250, main = "ACF of Amazon Log Returns")
pacf(amzn_sqr, lag.max = 100, main = "PACF of Amazon Log Returns")
```





# Model