---
title: "Kalman Filter Training"
output: pdf_document
---

```{r, echo = F, message = F, warning = F}
library(forecast)
```


# Basic Kalman Filtering

## Time series Setup

First, lets create some time series Data. Its verifiable that it is a short time series by the use of the ACF which shows a quick decay.

\begin{align*}
    x_t &= \phi x_{t-1} + w_t \\
    \phi &= 0.6, w_t \sim N(0,5)
\end{align*}

```{r, echo = F, message = F, warning = F}
set.seed(3476239)
mu0 <- 0
phi <- 0.6
Q <- sqrt(5)  # sd of white noise

X <- numeric(100)
X[1] <- rnorm(1, mu0, Q)
for (i in 2:100) {
  wt <- rnorm(1, 0, Q)
  X[i] <- phi * X[i - 1] + wt
}
par(mfrow=c(2,2))
plot(X, type = "l", main="Stationary AR(1) with mean 15, phi=0.6, sd=5")
abline(h = mu0, col = "red", lty = 2)
acf(X)
pacf(X)
```
\pagebreak

## AR(1) Walkthrough
Now we take a quick peek at the AR(1) Example of the data

```{r, echo = F, message = F, warning = F}
par(mfrow=c(1,2))
ARModel <- arima(X, order = c(1,0,0))
fitted_vals = X - ARModel$residuals
plot(X, type = "l", col = "black", main = "AR(1) Fit vs Actual")
lines(fitted_vals, col = "blue", lwd = 2)
legend("topleft", legend = c("Actual", "Fitted"), 
       col = c("black", "blue"), lty = 1, lwd = 2)
plot(ARModel$residuals, ylab="AR(1) Residuals", main="AR(1) Residual Plot")
```
\pagebreak 

## Kalman Filtering

Now onto the Kalman filter itself! This is best explained by the parameters K and P. If K and P are big, you rely heavily on the data. As it goes on, the P and K will decrease and smooth it out. Here's some math

Prediction step. let t-1 be the most recent observed value we have, and we want to predict the value of x at t
\begin{align*}
    \hat{X}_{t|t-1} &= \phi \hat{x}_{t-1|t-1}\\
    P_{t|t-1} &= \phi^2P_{t-1|t-1} + Q
\end{align*}

Ok, so you did the prediction, some time passed and you acquired the true value. What to do next? We update based on the Kalman Gain, which is a parameter that basically asks, "How good did we do?" and allows you to change your weight of reliance on prior information

\begin{align*}
    K_t &= \frac{P_{t|t-1}}{(P_{t|t-1} + R)}\\
    \hat{x}_{t|t} &= \hat{x}_{t|t-1} + K_t(y_t-\hat{x}_{t|t-1})\\
    P_{t|t} &= (1-K)P_{t|t-1}
\end{align*}


```{r, echo = F, message = F, warning = F}
phi <- 7
Q <- 5     # process noise variance
R <- 2     # observation noise variance (tune this), 2 works pretty well
n <- length(X)

xhat <- numeric(n)
P <- numeric(n)

# Initial guesses
xhat[1] <- X[1]
P[1] <- 1
K = c()

for (t in 2:n) {
  # Predict
  x_pred <- phi * xhat[t - 1]
  P_pred <- phi^2 * P[t - 1] + Q
  
  # Update
  k <- P_pred / (P_pred + R) # Kalman Gain, very important
  xhat[t] <- x_pred + k * (X[t] - x_pred)
  P[t] <- (1 - k) * P_pred
  K = c(K, k)
}

# Compare results
plot(X, type = "l", col = "gray", lwd = 2, main = "Kalman Filter vs Observations")
lines(xhat, col = "blue", lwd = 2)
legend("bottomright", legend = c("Observed X", "Filtered estimate"), 
       col = c("gray", "blue"), lty = 1, lwd = 2)


```


## Kalman Smoothing 

```{r}
phi <- 0.6
Q <- 5
R <- 2
n <- length(X)

# --- Forward pass (filtering) ---
xhat <- numeric(n)
P <- numeric(n)
x_pred <- numeric(n)
P_pred <- numeric(n)

xhat[1] <- X[1]
P[1] <- 1

for (t in 2:n) {
  # Predict
  x_pred[t] <- phi * xhat[t - 1]
  P_pred[t] <- phi^2 * P[t - 1] + Q
  
  # Update
  K <- P_pred[t] / (P_pred[t] + R)
  xhat[t] <- x_pred[t] + K * (X[t] - x_pred[t])
  P[t] <- (1 - K) * P_pred[t]
}

# --- Backward pass (smoothing) ---
x_smooth <- xhat
P_smooth <- P

for (t in (n - 1):1) {
  A <- P[t] * phi / P_pred[t + 1]
  x_smooth[t] <- xhat[t] + A * (x_smooth[t + 1] - x_pred[t + 1])
  P_smooth[t] <- P[t] + A^2 * (P_smooth[t + 1] - P_pred[t + 1])
}

# --- Plot results ---
plot(X, type = "l", col = "gray", lwd = 2, main = "Kalman Filter & Smoother")
lines(xhat, col = "blue", lwd = 2)
lines(x_smooth, col = "red", lwd = 2)
legend("topleft", legend = c("Observed", "Filtered", "Smoothed"),
       col = c("gray", "blue", "red"), lty = 1, lwd = 2)

```

# ARMA(1,2) Example

## Data Creation

This time we forego the independent creation of data and use a conventional one. ARIMA.sim(2,2,3) to create a more complex situation

```{r, echo = F, message = F, warning = F}
set.seed(897)

# ---------- params ----------
n <- 300
phi    <- 0.7
theta1 <- 0.4
theta2 <- -0.25
sigma2 <- 1    # innovation variance


# ---------- simulate ARMA(1,2) ----------
z <- rnorm(n+50, 0, sqrt(sigma2))  # extras for burn-in
x <- numeric(length(z))
for (t in 3:length(z)) {
  x[t] <- phi * x[t-1] +
          theta1 * z[t-1] + theta2 * z[t-2] +
          z[t]
}
x <- x[-(1:50)]   # drop burn-in
z <- z[-(1:50)]
tsdata <- as.numeric(x)

# par(mfrow=c(1,3))
plot(tsdata, type = "l", col = "steelblue",
     main = "Simulated ARIMA(2,2,3)",
     ylab = "Value", xlab = "Time")
acf(tsdata)
pacf(tsdata)
```

## Fitting using the model

Spot on! Good work! Our ARMA(1,2) model perfectly fits out simulated ARMA(1,2) data!

```{r}
ARMAModel <- arima(tsdata, order = c(1,0,2))
ARMAModel
par(mfrow=c(1,2))
MSE = mean(ARMAModel$residuals^2)
print(paste("MSE of model is:", MSE))
fitted_vals = fitted(ARMAModel)
plot(tsdata, type = "l", col = "black", main = "ARIMA(2,2,3) Fit vs Actual")
lines(fitted_vals, col = "blue", lwd = 2)
legend("topleft", legend = c("Actual", "Fitted"), 
       col = c("black", "blue"), lty = 1, lwd = 2)
plot(ARMAModel$residuals, ylab="ARIMA(2,2,3) Residuals", main="AR(1) Residual Plot")
```

## And for the kalman filter

```{r}
# ---------- params ----------
n <- 300
phi    <- 0.7
theta1 <- 0.4
theta2 <- -0.25
sigma2 <- 1    # innovation variance
Robs   <- 0.0001  # observation noise variance (tiny; set 0 if you want exact obs)


# ---------- state-space matrices ----------
F <- matrix(c(phi, theta1, theta2,
              0,   0,      0,
              0,   1,      0), nrow=3, byrow=TRUE)
B <- matrix(c(1, 1, 0), nrow=3)
H <- matrix(c(1,0,0), nrow=1)
Q <- sigma2 * (B %*% t(B))
R <- Robs

# ---------- allocate storage ----------
Tn <- length(tsdata)
m <- nrow(F)
x_pred <- matrix(0, m, Tn)   # x_{t|t-1}
P_pred <- array(0, c(m,m,Tn))
x_filt <- matrix(0, m, Tn)   # x_{t|t}
P_filt <- array(0, c(m,m,Tn))

# ---------- init ----------
x_filt[,1] <- c(tsdata[1], 0, 0)   # initial guess (observed first value)
P_filt[,,1] <- diag(100, m)        # large uncertainty to start

# ---------- forward Kalman filter ----------
for (t in 2:Tn) {
  # predict
  x_pred[,t] <- F %*% x_filt[,t-1]
  P_pred[,,t] <- F %*% P_filt[,,t-1] %*% t(F) + Q

  # update using y_t
  S <- H %*% P_pred[,,t] %*% t(H) + R
  K <- P_pred[,,t] %*% t(H) %*% solve(S)   # Kalman gain (m x 1)
  innov <- tsdata[t] - (H %*% x_pred[,t])
  x_filt[,t] <- x_pred[,t] + K %*% innov
  P_filt[,,t] <- (diag(m) - K %*% H) %*% P_pred[,,t]
}

# ---------- Rauch-Tung-Striebel smoother (backwards) ----------
x_smooth <- x_filt
P_smooth <- P_filt
for (t in (Tn-1):1) {
  C <- P_filt[,,t] %*% t(F) %*% solve(P_pred[,,t+1])
  x_smooth[,t] <- x_filt[,t] + C %*% (x_smooth[,t+1] - x_pred[,t+1])
  P_smooth[,,t] <- P_filt[,,t] + C %*% (P_smooth[,,t+1] - P_pred[,,t+1]) %*% t(C)
}

# ---------- extract */
x_filtered <- as.numeric(x_filt[1,])
x_smoothed <- as.numeric(x_smooth[1,])

# ---------- plots ----------
plot(tsdata, type="l", main="ARMA(1,2): observed vs filtered & smoothed", ylab="x", xlab="t")
lines(x_filtered, lty=2, lwd=2)
lines(x_smoothed, col="red", lwd=2)
legend("topleft", legend=c("observed","filtered","smoothed"), lty=c(1,2,1), col=c("black","black","red"))


MSE = mean((tsdata-x_filtered)^2)
print(paste("MSE of model is:", MSE))
```


# ARTFIMA Example
