---
title: "Single Location Expanded"
output: pdf_document
---

```{r}
library(artfima)
source("ArtfimaCausal.R")
source("StateSpaceSteps.R")
```


So after reading 4. Simulation Studies, I discovered that I am not quite approaching this the right way. Half the point of this paper is sneakily breezed through in like 3 paragraphs. 

So, heres my work flow

  + We start with true parameters and simulate T observations at N locations.
  + We then treat this simulated data as observed data.
  + To estimate parameters, we start with initial guesses, build the state-space matrices, and run the Kalman filter to compute the Gaussian likelihood.
  + An optimizer updates the parameters to maximize this likelihood, repeatedly running the Kalman filter until convergence.
  + This whole process is repeated M times to study finite-sample behavior.

Lets do a walk through of an ARMA(1,1) just like they did

# True Data Creation
```{r}
set.seed(4545)
true_theta = list(
  d = 0, # fractional differencing
  lambda = 0, # tempered
  phi = 0.45, # AR
  theta = 0.35, # MA
  sigma2_eta = 1.3, # Temporal Noise
  sigma2_w = 0.58 # Measure Noise
)
# data params
T = 100
N=1
m = 10

A = artsim(n=T, d=true_theta$d, lambda=true_theta$lambda,
           phi=true_theta$phi, theta=true_theta$theta, sigma2 = true_theta$sigma2_eta)

A = A + rnorm(T, 0, sqrt(true_theta$sigma2_w))
mat = matrix(A, nrow=N, ncol=T,
             byrow=TRUE)
```

# Implicit calculations

## initial guesses
```{r}
# Guesses
dhat = 0
lambdahat = 0
phihat = 0.75
thetahat = 0.75
sigma2_eta = 0.75
sigma2_w = 0.2
psi <- psi_artfima(
  m       = m,
  d       = dhat,
  lambda  = lambdahat,
  ar      = phihat,
  ma      = thetahat
)

#equivalent, psi has an intercept term though.
ARMAtoMA(ar=phihat, ma = thetahat, m)
psi
```

## State space creation!

```{r}
# Dimensions
n_loc <- nrow(mat)
p <- 1
state_dim = m + 1


# Observation noise
#TODO: this is a possible contribution I could add!!! Fereirra assumes this is 0 actually
R <- diag(sigma2_w, n_loc)


# Observation equation: Y_t = G*X_t + M*beta + W_t
G <- matrix(psi[1:(m+1)], nrow = n_loc, ncol = m + 1)  # [psi_0, psi_1, ..., psi_m]
M <- matrix(1, nrow = n_loc, ncol = p) # Design matrix for mean
beta <- matrix(0, nrow = p, ncol = 1) # TODO estimate this later maybe?


# State equation: X_t = F*X_{t-1} + H*eta_t
F <- matrix(0, state_dim, state_dim)
F[2:state_dim, 1:(state_dim - 1)] <- diag(state_dim - 1)

H <- matrix(0, state_dim, n_loc)
H[1:n_loc, 1:n_loc] <- diag(n_loc)
```

## Running 

```{r}
# Initial Vectors
Y_hat = matrix(0, nrow=nrow(mat), ncol=ncol(mat)+1)
likelihood = 0

X_t <- matrix(0, nrow = m + 1, ncol = 1)
Omega_t <- diag(1000, m + 1)


for (i in 1:ncol(mat)) {
  # given previous instances predict.
  pred = predict_step(X_t, Omega_t, F, H, sigma2_eta, G, M, beta)
  # update values
  Y_hat[1,i] = pred$Y_pred
  X_t = pred$X_pred
  Omega_t = pred$Omega_pred
  
  # zzzz some time passes
  
  # Now you see the thing, so update our values
  Y_obs = mat[1,i]
  update = update_step(pred$X_pred, pred$Omega_pred, Y_obs, 
                        F, H, sigma2_eta, G, M, beta, sigma2_w)
  # Update values
  X_t = update$X_updated
  Omega_t = update$Omega_updated
  likelihood= likelihood + update$likeli
  
}

pred = predict_step(X_t, Omega_t, F, H, sigma2_eta, n_loc)
# update values
Y_hat[1,i+1] = pred$Y_pred
X_t = pred$X_pred
Omega_t = pred$Omega_pred
likelihood

# Sigma_Eta is the noise in the inherent state, directly related to X
# sigma_w is the measurement noise. Y = fX + sigma_w kinda thing


```


# State Space as a loop

## state space as functions

```{r}
create_state_space = function(n_loc, p, state_dim, sigma2_w, psi, m) {
  
  
  # Measurement Variance
  R <- diag(sigma2_w, n_loc)# measurement error variance
  
  # Observation matrix
  G <- matrix(psi[1:(m+1)], nrow = n_loc, ncol = m + 1)  # [psi_0, psi_1, ..., psi_m]
  
    # Mean structure
  # TODO estimate this or something?
  # Observation equation: Y_t = G*X_t + M*beta + W_t
  M <- matrix(1, nrow = n_loc, ncol = 1)
  beta <- matrix(0, nrow = p, ncol = 1)
  
  # Transition matrix
  # State equation: X_t = F*X_{t-1} + H*eta_t
  F <- matrix(0, state_dim, state_dim)
  F[2:state_dim, 1:(state_dim - 1)] <- diag(state_dim - 1)
  
    # Innovation injection
  H <- matrix(0, state_dim, n_loc)
  H[1:n_loc, 1:n_loc] <- diag(n_loc)
  
  return(list(
    G = G,
    M = M,
    beta = beta,
    F = F,
    H = H,
    R= R
  ))
}

kalman_filter = function(data, state_space, theta, m) {
  likelihood = 0
  n_loc = nrow(data)
  X_t = matrix(0, nrow = m + 1, ncol = nrow(data)) # this should be m+1 x n_loc basically. 
  Omega_t <- diag(500, m + 1)
  for (i in 1:ncol(data)) {
    pred = predict_step(X_t, Omega_t, state_space$F,
                        state_space$H, theta$sigma2_eta, n_loc)
    
    X_t = pred$X_pred
    Omega_t = pred$Omega_pred
    
    # zzz some time passes
    Y_obs = data[1,i]
    
    # update = update_step(pred$X_pred, pred$Omega_pred, Y_obs, 
    #                   state_space$F, state_space$H, theta$sigma2_eta,
    #                   state_space$G, state_space$M, state_space$beta,
    #                   theta$sigma2_w)
    
    update = update_step(pred$X_pred, pred$Omega_pred, Y_obs,
                        state_space$G, state_space$M,
                        state_space$beta, theta$sigma2_w)
    
    # Update values
    X_t = update$X_updated
    Omega_t = update$Omega_updated
    likelihood= likelihood + update$likeli
    
  }
  return(as.numeric(likelihood))
}


```

## Testing my functions

```{r}
# Initialization, this is all I have 
theta = list(
  dhat = 0,
  lambdahat = 0,
  phihat = 0.75,
  thetahat = 0.75,
  sigma2_eta = 0.75,
  sigma2_w = 0.2
)
m =10
psi <- psi_artfima(
  m       = m,
  d       = theta$dhat,
  lambda  = theta$lambdahat,
  ar      = theta$phihat,
  ma      = theta$thetahat
)
# mat and N x T matrix of observations!
statespace = create_state_space(n_loc = nrow(mat), p=1, state_dim = m +1, sigma2_w = theta$sigma2_w, psi=psi, m=m)
likeli = kalman_filter(mat, statespace, theta, m)
```

## wrapped in a big function
```{r}
neg_loglik <- function(par, data, m) {
  
  theta <- list(
    dhat = if ("dhat" %in% names(par)) par[['dhat']] else 0,
    lambdahat = if ("lambdahat" %in% names(par)) par[['lambdahat']] else 0,
    phihat = if ("phihat" %in% names(par)) par[['phihat']] else 0,
    thetahat = if ("thetahat" %in% names(par)) par[['thetahat']] else 0,
    sigma2_eta = if ("sigma2_eta" %in% names(par)) par[['sigma2_eta']] else 0,
    sigma2_w = if ("sigma2_w" %in% names(par)) par[['sigma2_w']] else 0
  )
  
  if (theta$sigma2_eta <= 0 || theta$sigma2_w <= 0) {
    return(1e12)
  }

  psi <- psi_artfima(
    m = m,
    d = theta$dhat,
    lambda = theta$lambdahat,
    ar = theta$phihat,
    ma = theta$thetahat
  )

  statespace <- create_state_space(
    n_loc = nrow(data),
    p = 1,
    state_dim = m+1,
    sigma2_w = theta$sigma2_w,
    psi = psi,
    m = m
  )

  lik <- kalman_filter(data, statespace, theta, m)
  
  return(as.numeric(-lik))

    # return negative log likelihood
}
```

## And run?

```{r}
thetahat = list(
  # dhat = 0,
  # lambdahat = 0,
  phihat = 0.5,
  thetahat = 0.5,
  sigma2_eta = 1,
  sigma2_w = 0.2
)

# 0 indicates successful convergence
fit = nlminb(
  start = unlist(thetahat),
  objective = neg_loglik,
  data = mat,
  m = 10,
  lower = c(-0.99, -0.99, 1e-6, 1e-6),
  upper = c(0.99, 0.99, 10, 10)
)

unlist(true_theta)
fit
```

## Monte Carlo Experiment

To be run when my stuff stops sucking
```{r}
true_theta = list(
  d = 0,
  lambda = 0,
  phi = 0.45,
  theta = 0.35,
  sigma2_eta = 1.3,
  sigma2_w = 0.1
)

#initial guesses
thetahat = list(
  # dhat = 0,
  # lambdahat = 0,
  phihat = 0.75,
  thetahat = 0.75,
  sigma2_eta = 0.75,
  sigma2_w = 0.2
)

# data params
T = 100
N=1
m = 10

iter= 5

estimates = matrix(NA, nrow=iter, ncol=length(thetahat))
colnames(estimates) = names(thetahat)

for (i in 1:iter) {
  # Data creation
  A = artsim(n=T, d=true_theta$d, lambda=true_theta$lambda,
           phi=true_theta$phi, theta=true_theta$theta, sigma2 = true_theta$sigma2_eta)

  A = A + rnorm(T, 0, sqrt(true_theta$sigma2_w))
  mat = matrix(A, nrow=N, ncol=T,
               byrow=TRUE)
  

  
  # estimation
  fit = nlminb(
    start = thetahat,
    objective = neg_loglik,
    data = mat,
    m = 10,
    lower = c(-0.99, -0.99, 1e-6, 1e-6),
    upper = c(0.99, 0.99, 10, 10)
  )
  # collect
  estimates[i,] = fit$par
  
}

true_vec = unlist(true_theta)[3:6]
# Mean estimate
mean_est = colMeans(estimates, na.rm = TRUE)

# Bias
bias = mean_est - true_vec

# Relative Bias
rel_bias = bias / true_vec

# MSE
mse = colMeans((estimates - matrix(true_vec, 
                                    nrow=iter, 
                                    ncol=length(true_vec),
                                    byrow=TRUE))^2,
               na.rm=TRUE)

results = rbind(
  True = true_vec,
  Mean = mean_est,
  Bias = bias,
  RelBias = rel_bias,
  MSE = mse
)

print(results)

```

