---
title: "Multi Location Expanded"
output: pdf_document
---

# Introduction

```{r, warning=FALSE}
library(artfima)
library(MASS)
library(parallel)

source("ArtfimaCausal.R")
source("MultiStateSpaceFunctions.R")
source("Data_Creation_Funcs.R")
```


So after reading 4. Simulation Studies, I discovered that I am not quite approaching this the right way. Half the point of this paper is sneakily breezed through in like 3 paragraphs. 

So, heres my work flow

  + We start with true parameters and simulate T observations at N locations.
  + We then treat this simulated data as observed data.
  + To estimate parameters, we start with initial guesses, build the state-space matrices, and run the Kalman filter to compute the Gaussian likelihood.
  + An optimizer updates the parameters to maximize this likelihood, repeatedly running the Kalman filter until convergence.
  + This whole process is repeated M times to study finite-sample behavior.

Lets do a walk through of an ARMA(1,1) just like they did

# True Data Creation
```{r}
set.seed(4545)

true_theta <- list(
  d = 0,
  lambda = 0,
  phi = 0.35,
  theta = 0.35,
  sigma2_eta = 3,
  sigma2_w = 0,
  rho = 0.7
)

# -------------------------
# Dimensions
# -------------------------
T_len <- 150
n_side <- 3
N <- n_side^2
m <- 10

# -------------------------
# 1️⃣ Build spatial grid
# -------------------------
coords <- expand.grid(x = 1:n_side, y = 1:n_side)
coords <- as.matrix(coords)

# Pairwise distances
D <- as.matrix(dist(coords))

# -------------------------
# 2️⃣ Spatial innovation covariance (Matérn ν = 1/2)
# -------------------------
Sigma_eta <- true_theta$sigma2_eta * exp(-D / true_theta$rho)

# -------------------------
# 3️⃣ Temporal filter weights (ψ)
# -------------------------
psi <- psi_artfima(
  m = m,
  d = true_theta$d,
  lambda = true_theta$lambda,
  ar = true_theta$phi,
  ma = true_theta$theta
)

# -------------------------
# 4️⃣ Generate spatial innovations η_t  (N × T)
# -------------------------

eta <- matrix(0, N, T_len + m)

for (t in 1:(T_len + m)) {
  eta[, t] <- mvrnorm(1, mu = rep(0, N), Sigma = Sigma_eta)
}

# -------------------------
# 5️⃣ Build latent process X_t via convolution
# -------------------------
X <- matrix(0, N, T_len)

for (t in 1:T_len) {
  for (j in 0:m) {
    X[, t] <- X[, t] + psi[j + 1] * eta[, t + m - j]
  }
}

# -------------------------
# 6️⃣ Add measurement noise
# -------------------------
W <- matrix(rnorm(N * T_len, 0, sqrt(true_theta$sigma2_w)),
            nrow = N, ncol = T_len)

Y <- X + W

# Final data matrix
mat <- Y
```

# First Smoke Test

```{r}
# -------------------------
# Smoke test: single likelihood evaluation
# -------------------------

m <- 10
n_loc <- nrow(mat)
n_side <- 3  # since 3x3 grid

# Parameter guess
thetahat <- list(
  dhat = 0,
  lambdahat = 0.4,
  phihat = 0.5,
  thetahat = 0.75,
  sigma2_eta = 0.75,
  sigma2_w = 0.2,
  rho = 0.5
)

# -------------------------
# 1️⃣ Build psi
# -------------------------
psi <- psi_artfima(
  m = m,
  d = thetahat$dhat,
  lambda = thetahat$lambdahat,
  ar = thetahat$phihat,
  ma = thetahat$thetahat
)

# -------------------------
# 2️⃣ Build spatial covariance (Matérn ν = 1/2)
# -------------------------
coords <- expand.grid(x = 1:n_side, y = 1:n_side)
coords <- as.matrix(coords)

D <- as.matrix(dist(coords))

Sigma_eta_hat <- thetahat$sigma2_eta *
  exp(-D / thetahat$rho)

# -------------------------
# 3️⃣ Build state-space
# -------------------------
state_space <- create_state_space(
  n_loc = n_loc,
  m = m,
  sigma2_w = thetahat$sigma2_w,
  psi = psi,
  Sigma_eta = Sigma_eta_hat
)

# -------------------------
# 4️⃣ Run Kalman filter (Expanding to debug)
# -------------------------
loglik_value <- kalman_filter(
  data = mat,
  state_space = state_space,
  m = m
)

loglik_value
```



## And run?

```{r}
thetahat <- c(
  phihat = 0.5,
  thetahat = 0.5,
  sigma2_eta = 1,
  rho = 0.5
)

fit <- nlminb(
  start = thetahat,
  objective = neg_loglik,
  data = mat,
  m = 10,
  n_side = 3,
  # phihat, thetahat, sigma2_eta, sigma2_w, rho
  lower = c(-0.99, -0.99, 1e-6, 1e-6, 1e-6),
  upper = c(0.99, 0.99, 10, 10, 10)
)

unlist(true_theta)
fit
```

## Monte Carlo Experiment

### Parallelize run

Parallelize prep

```{r}
mc_one_run <- function(i, true_theta, start_vals, T_len, n_side, m, lower, upper) {
  
  # Simulate
  mat <- simulate_artfima_spatial(
    T_len = T_len,
    n_side = n_side,
    m = m,
    true_theta = true_theta
  )
  
  # Estimate
  fit <- tryCatch(
    nlminb(
      start = start_vals,
      objective = neg_loglik,
      data = mat,
      m = m,
      n_side = n_side,
      lower = lower,
      upper = upper
    ),
    error = function(e) return(NULL)
  )
  
  if (!is.null(fit) && fit$convergence == 0) {
    return(fit$par)
  } else {
    return(rep(NA, length(start_vals)))
  }
}
```

```{r}
n_cores <- detectCores() - 1

set.seed(123)

# -------------------------
# True parameters
# -------------------------
true_theta <- list(
  d = 0.25,
  lambda = 0.01,
  phi = 0.35,
  theta = 0.45,
  sigma2_eta = 1.5,
  sigma2_w = 0,
  rho = 0.55
)

# -------------------------
# MC settings
# -------------------------
n_side <- 3
T_len <- 100
m <- 10
iter <- 15

# -------------------------
# Starting values
# -------------------------
start_vals <- c(
  dhat = 0.3,
  lambdahat = 0.1,
  phihat = 0.2,
  thetahat = 0.2,
  sigma2_eta = 1,
  rho = 0.25
)

# lower bounds 
lower = c(-0.99, 1e-6, -0.99, -0.99, 1e-6, 1e-6)
upper = c(0.99, 0.2, 0.99, 0.99, 10, 10)
```

```{r}
cl <- makeCluster(n_cores)

clusterExport(cl, ls())  # export all objects
clusterEvalQ(cl, library(MASS))  # needed for mvrnorm

results_list <- parLapply(
  cl,
  1:iter,
  mc_one_run,
  true_theta = true_theta,
  start_vals = start_vals,
  T_len = T_len,
  n_side = n_side,
  m = m,
  lower = lower,
  upper = upper
)

stopCluster(cl)

estimates <- do.call(rbind, results_list)
true_vec <- c(
  dhat = true_theta$d,
  lambdahat = true_theta$lambda,
  phihat = true_theta$phi,
  thetahat = true_theta$theta,
  sigma2_eta = true_theta$sigma2_eta,
  # sigma2_w = true_theta$sigma2_w,
  rho = true_theta$rho
)

mean_est <- colMeans(estimates, na.rm = TRUE)
sd_est <- apply(estimates, 2, sd, na.rm = TRUE)

bias <- mean_est - true_vec
rel_bias <- bias / true_vec

mse <- colMeans(
  (estimates -
     matrix(true_vec,
            nrow = iter,
            ncol = length(true_vec),
            byrow = TRUE))^2,
  na.rm = TRUE
)
SqrtMSE = sqrt(mse)

results <- rbind(
  True = true_vec,
  Mean = mean_est,
  sd = sd_est,
  RelBias = rel_bias,
  SqrtMSE = SqrtMSE
)

print(round(results,4))
# print(results)

```



### Mono-core run

To be run when my stuff stops sucking
```{r}
set.seed(123)

# -------------------------
# True parameters
# -------------------------
true_theta <- list(
  d = 0.15,
  lambda = 0,
  phi = 0,
  theta = 0.45,
  sigma2_eta = 1.5,
  sigma2_w = 0.1,
  rho = 0.55
)

# -------------------------
# MC settings
# -------------------------
n_side <- 3
T_len <- 100
m <- 10
iter <- 15

# -------------------------
# Starting values
# -------------------------
start_vals <- c(
  dhat = 0.3,
  thetahat = 0.2,
  sigma2_eta = 1,
  sigma2_w = 0.2,
  rho = 0.25
)

estimates <- matrix(NA, nrow = iter, ncol = length(start_vals))
colnames(estimates) <- names(start_vals)

# -------------------------
# Monte Carlo Loop
# -------------------------
for (i in 1:iter) {
  
  cat("Iteration:", i, "\n")
  
  # ---- Simulate spatial data ----
  mat <- simulate_artfima_spatial(
    T_len = T_len,
    n_side = n_side,
    m = m,
    true_theta = true_theta
  )
  
  # ---- Estimate parameters ----
  fit <- nlminb(
    start = start_vals,
    objective = neg_loglik,
    data = mat,
    m = m,
    n_side = n_side,
    lower = c(-0.99, -0.99, 1e-6, 1e-6, 1e-6),
    upper = c(0.99, 0.99, 10, 10, 10)
  )
  
  if (fit$convergence == 0) {
    estimates[i, ] <- fit$par
  } else {
    cat("Non-convergence in iteration", i, "\n")
  }
}

# -------------------------
# Monte Carlo Summary
# -------------------------

true_vec <- c(
  # phihat = true_theta$phi,
  dhat = true_theta$d,
  thetahat = true_theta$theta,
  sigma2_eta = true_theta$sigma2_eta,
  sigma2_w = true_theta$sigma2_w,
  rho = true_theta$rho
)

mean_est <- colMeans(estimates, na.rm = TRUE)
bias <- mean_est - true_vec
rel_bias <- bias / true_vec

mse <- colMeans(
  (estimates -
     matrix(true_vec,
            nrow = iter,
            ncol = length(true_vec),
            byrow = TRUE))^2,
  na.rm = TRUE
)

results <- rbind(
  True = true_vec,
  Mean = mean_est,
  Bias = bias,
  RelBias = rel_bias,
  MSE = mse
)

# print(round(results,4))
print(results)
```

